% Save file as: APP-G.TEX              Source: FILESERV@SHSU.BITNET  
\input fontsize.tex
\input g.tex
\nopagenumbers
\headline={\hfill}
\twelvepoint
\centerline{\twlbf APPENDIX G: Incorporation of PostScript Graphics}
\bigskip\par\noindent
So far you've heard no mention of graphics in conjunction with \TeX{} --- 
this is because \TeX\ itself makes no explicit provision for their 
inclusion!
\bigskip\par\noindent
\TeX\ {\twlit does,} however, provide a ``{\twltt{}$\backslash$special}''
command which implements ``user-exits'' (or ``hooks'') which can be 
exploited by an intelligent DVI-to-$<$whatever$>$ postprocessor to extend
\TeX{}. One such extension provided by the DVI-to-PostScript converter
{\twltt DVIPS} is inclusion of PostScript graphics files.
\bigskip\par\noindent
NOTE: Do not embark upon including PostScript files in your \TeX\ document 
lightly; including graphics this way is a big decision for a 
couple of reasons:
\bigskip\par\noindent
(1) The quality of the PostScript code produced by applications can
vary immensely. Some nominal ``PostScript'' output can be so ill-formed 
you'll find yourself wondering how {\twlit any} PostScript device has 
every been able to interpret and print it. This ``low-grade'' PostScript
may cause all sorts of headaches when you try to include it in your
document.
\bigskip\par\noindent
(2) If you do elect to include PostScript graphics in your \TeX\ document,
you'll find that virtually no DVI previewer will be able to show you your 
graphic inclusions on screen during the output previewing process; you'll
need to use a PostScript previewer (like NeXT's Preview application) 
if you need to look at your output on screen before actually printing it.
\bigskip\par\noindent
(3) Inclusion of PostScript graphics in your document excludes, virtually
by definition, all DVI-to-$<$whatever$>$ converters other than DVIPS. Are
you {\twlit sure} you'll never need to do output to a non-PostScript output
device, such as an HP LaserJet printer?
\bigskip\par\noindent
(4) Even small illustrations tend to generate huge PostScript files.
Do you have the disk space you'll need to work with the images you'll be
generating?
\bigskip\par\noindent
Nonetheless, if in spite of all these potential problems you decide that 
you really {\twlit do} want to include PostScript graphics in your document, 
take a look at the code which was used to generate this appendix, and be
sure to give yourself {\twlit lots} of extra time to overcome any problems
you may encounter in working with this feature.
\bigskip\par\noindent
A machine readable copy of the raw \TeX\ commands used 
is available to OREGON VAX users as:
\bigskip\par\noindent\hglue 0.5truein
{\twltt TEX\$DISK:[TEX.SAMPLES]EPSGRAPHS.TEX}
\bigskip\par\noindent
On other systems get a copy of that file via anonymous FTP. 
\bigskip\par\noindent
FTP to {\twltt DECOY.UOREGON.EDU (128.223.32.19)} and see directory 
{\twltt pub/tex/samples}.
\bigskip\par\noindent
\vfill\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input epsgraphs.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end
