\nopagenumbers
\headline={\hfill}
\hsize=6.0truein
\vsize=8.75truein
\hoffset=0.5truein
\voffset=0.25truein
\input fontsize.tex
\twelvepoint
\settabs 12 \columns % establish a dozen tab stops
\def\leaderfill{\leaders\hbox to 0.4em{\hss.\hss}\hfil}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% NOTE: FOLLOWING PAGE ORDER IS PER PAGE 5 OF THE U OF O STYLE MANUAL
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% title page (see pp. 27 of the U of O Style Manual) %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vglue .75truein      % .75"+1.25" (margin)=2" per pp. 25 of U of O Style Manual
\centerline{ANALYSIS OF UNUSUAL CONFORMATIONS OF OLIGIO}
\bigskip
\centerline{AND POLYNUCLEOTIDES AND PEPTIDES}
\bigskip
\centerline{BY CLASSIC RAMAN SCATTERING}
\vglue 2.50truein   % per pp. 25 (shrunk .50" to fit the page)
\centerline{by}
\bigskip
\centerline{GERALD ALAN THOMAS}
\vglue 2.25truein   % per pp. 25 (shrunk .25" to fit the page)
\centerline{A DISSERTATION}
\bigskip
\centerline{Presented to the Department of Chemistry}
\centerline{and the Graduate School of the University of Oregon}
\centerline{in partial fulfillment of the requirements}
\centerline{for the degree of}
\centerline{Doctor of Philosophy}
\bigskip
\centerline{August 1984}
\vfill
\eject
\pageno=-2      % use lowercase roman numerals for prefatory material
\headline={\tenrm\hfil\folio\voffset=2\normalbaselineskip}     
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% approval page (see pp. 10 of the U of O Style Manual %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vglue 2.75truein
\line{\hfill APPROVED: {\vbox{\hrule width 3 truein height 0.5pt}} \hfill}
\+&&&&&{Dr. Harry S. Truman}\cr
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% copyright page (see pp. 10 of the U of O Style Manual %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vglue 2.75truein
\line{\hfill \copyright\/\/ 1984 Jane Doe \hfill}
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% abstract page (see pp. 29 of the U of O Style Manual %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\baselineskip=2\normalbaselineskip      % double space it...
\vglue 0.75truein
\centerline{An Abstract of the Dissertation of}
\line{Walter James Teasdale\hfil for the degree of\hfil Doctor of Philosophy}
\line{in the Department of Economics\hfil to be taken\hfil August 1984}
\+Title:&{ADOPTION AND IMPLEMENTATION OF THE NORTHWEST POWER}\cr
\+&{PLANNING COUNCIL'S MODEL CONSERVATION STANDARDS}\cr
\bigskip\bigskip\bigskip\baselineskip=\normalbaselineskip  
\+Approved: {\vbox{\hrule width 3truein height 0.5pt}}\cr
\+&&&{Dr. Harry S. Truman}\cr
\baselineskip=2\normalbaselineskip
\bigskip\bigskip		 
The Pacific Northwest Electric Power Planning and Conservation Act
of 1980 and the Northwest Conservation and Power Plan are models for 
economic planning. It is an essential part of the Council's resource
portfolio. The Standards set energy use budgets for new buildings and 
(text continues)
%%% NOTE: 350 word maximum for doctoral dissertation abstracts (150 word
%%% maximum for master's thesis abstracts)!
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% vita page (see pp. 33 of the U of O Style Manual %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\baselineskip=\normalbaselineskip
\vglue .75truein
\centerline{VITA}
\bigskip
\+NAME OF AUTHOR:\quad Jane Ann Doe\cr
\bigskip
\+PLACE OF BIRTH:\quad Portland, Oregon\cr   %%% optional
\bigskip
\+DATE OF BIRTH:\quad January 1, 1950\cr     %%% optional
\bigskip
\bigskip
\+GRADUATE AND UNDERGRADUATE SCHOOLS ATTENDED:\cr
\medskip
\+&University of Oregon\cr
\+&San Francisco State College\cr
\+&San Jose State College\cr
\bigskip
\bigskip		 
\+DEGREES AWARDED:\cr
\medskip
\+&Doctor of Philosophy, 1984, University of Oregon\cr
\+&Master of Arts, 1972, San Francisco State College\cr
\+&Bachelor of Arts, 1970, San Jose State College\cr
\bigskip
\bigskip		 
\+AREAS OF SPECIAL INTEREST:\cr
\medskip
\+&Psychology of Perception\cr
\bigskip
\bigskip		 
\+PROFESSIONAL EXPERIENCE:\cr
\medskip
\+&Research Assistant, Department of Psychology, Univeristy of\cr
\+&\quad Oregon, Eugene, 1973-75\cr
\medskip	   
\+&Teaching Assistant, Department of Psychology, University of\cr
\+&\quad Oregon, Eugene, 1972-73\cr
\bigskip
\bigskip		   
\+AWARDS AND HONORS:\cr
\medskip
\+&Danforth Fellowship, 1976-77\cr
\bigskip
\bigskip		 
\+PUBLICATIONS:\cr
\medskip
\+&(if any publications, list here, using same style sheet you use for\cr
\+&your bibliography; if none, omit this heading)\cr
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% acknowledgment page -- optional (see pp. 34 of the U of O Style Manual) %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\baselineskip=2\normalbaselineskip
\vglue .75truein
\centerline{ACKNOWLEDGMENTS}
\bigskip
The author expresses sincere appreciation to Professors Smith and Jones for 
their assistance in the preparation of this manuscript. In addition, special
thanks are due to Mr. John Doe, whose familiarity with the needs and ideas of
the community were helpful during the early programming phase of this undertaking.
I also thank the members of the community councils for their valuable input. 
The investigation was supported in part by a Public Health Service Predoctoral
Fellowship, Number 4 FL MH-21, 824-02, from the National Institutes of Mental
Health, and by a grant from the National Science Foundation, SPI 801-2372, to
Dr. Philip Smith of the University of Oregon.
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% dedication page -- optional page (see pp. 13 of the U of O Style Manual) %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\baselineskip=2\normalbaselineskip
\vglue 2.75truein
\centerline{DEDICATION}
\bigskip
\centerline{This manuscript is dedicated to my wife and children.}
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% table of contents page (see pp. 35 of the U of O Style Manual %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\baselineskip=\normalbaselineskip
\centerline{TABLE OF CONTENTS}
\bigskip
\halign{\hfil#&\quad# \leaderfill&\hfil#\cr
Chapter&\hglue 4.6truein&Page\cr
&\hfill&\cr
I.&INTRODUCTION&1\cr
\hglue .6truein&\hfill&\hglue .6truein\cr
II.&REVIEW OF LITERATURE&9\cr
&\hfill&\cr
&\quad Current Research on Marital Status and Income&9\cr
&\quad Research on Sex Roles&9\cr
&\quad The Relationship Between Marital Status and\hfill&\hfill\cr
&\quad \quad Income Following the Birth of the First Child&26\cr
&\quad Summary&34\cr
&\hfill&\cr
III.&METHODOLOGY&41\cr
&\hfill&\cr
&\quad Subjects&42\cr
&\quad Instruments&44\cr
&\quad General Procedures&53\cr
&\hfill&\cr
IV.&RESULTS&55\cr
&\hfill&\cr
&\quad Reality Hypothesis Result&56\cr
&\quad Tangent Hypothesis Result&57\cr
&\quad Circular Hypothesis Result&59\cr
&\quad Summary&62\cr
&\hfill&\cr
V.&SUMMARY OF FINDINGS AND CONCLUSIONS&67\cr
&\hfill&\cr
&\quad Major Findings&68\cr
&\quad Study Limitations&71\cr
&\quad Suggestions for Future Studies&81\cr
&\hfill&\cr}
\halign{#\hfil\cr
APPENDIX\hfill\cr}
\halign{\hfil#&\quad# \leaderfill&\hfil#\cr
\hglue 0.6truein&\hglue 4.6truein&\hglue 0.6truein\cr
A.&SUBJECT PROFILES&90\cr
&\hfill&\cr
B.&INTERVIEW QUESTIONNAIRE&98\cr
&\hfill&\cr
C.&LETTER TO PARTICIPANTS&99\cr
&\hfill&\cr
D.&OTHER DATA&100\cr}
\halign{#\quad\leaderfill&\hfil#\cr
\hglue 5.2truein&\hglue 0.6truein\cr
BIBLIOGRAPHY\quad&123\cr}
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% list of tables page (see pp. 36 of the U of O Style Manual %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\baselineskip=\normalbaselineskip
\vglue .75truein
\centerline{LIST OF TABLES}
\bigskip
\halign{\hfil#&\quad# \leaderfill&\hfil#\cr
Table&\hglue 4.6truein&Page\cr
&\hfill&\cr
1.&Frequencies and Percentages for Geographic Data\hfill&\cr
&\quad of Subjects' Marital Status and Average Age\hfill&\cr
&\quad as Categorized by Group and Income&77\cr
\hglue .6truein&\hfill&\hglue .6truein\cr
2.&Frequencies and Percentages for Geographic Data\hfill&\cr
&\quad of Subjects' Sex, Marital Status, and Average\hfill&\cr
&\quad Age Distributed Across Operational Systems&78\cr
\hglue .6truein&\hfill&\hglue .6truein\cr
3.&Mean, Standard Deviation, and Standard Error Scores\hfill&\cr
&\quad of Responses Integrating Input Compatibility&79\cr
\hglue .6truein&\hfill&\hglue .6truein\cr
4.&Summary Table for Frequencies and Percentages of\hfill&\cr
&\quad Responses Integrating Input Compatibility&80\cr
\hglue .6truein&\hfill&\hglue .6truein\cr
5.&Summary Table for Marital Status Contrasted\hfill&\cr
&\quad with Sequential Output Data&81\cr
\hglue .6truein&\hfill&\hglue .6truein\cr
6.&Summary Table for One-Way ANOVA Performed on\hfill&\cr
&\quad Multiple Interactive Compatibility Data&82\cr
\hglue .6truein&\hfill&\hglue .6truein\cr
7.&Summary Table for One-Way ANOVA Performed on\hfill&\cr
&\quad Functional Capabilities Data&83\cr}
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% text pages (see pp. 13-14 of the U of O Style Manual %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\pageno=1\baselineskip=2\normalbaselineskip
\vglue .75truein
\centerline{CHAPTER I}
\bigskip
\centerline{INTRODUCTION}
\bigskip
The spirit of this style manual is an inquiry into the parameters
set by necessity in order to standardize publications from graduate
students at the University of Oregon. The \underbar{Manual} is to be 
used as a guide. It does not contain all requirements. Other style
manuals should be considered.
\bigskip
\centerline{\underbar{Style Manuals}}
\bigskip
Other manuals are available at the University Bookstore for information.
Since the student has to make a choice concerning style of the manuscript
at an early date, we suggest using the library before deciding which style manual
to purchase. The Graduate School suggests using the manuals published by 
the University of Chicago.
\bigskip
\centerline{Chicago Style Manuals}
\bigskip
There are two manuals available from the University of Chicago
which the Graduate School uses in making final determination on questions
not addressed in the Graduate School Style Manual.
\vfill\eject     % added to prevent a widowed section heading at end of page
\noindent\underbar{Turabian}\hfill
\bigskip
The \underbar{Manual for Writers of Term Papers, Theses, and
Dissertations} (Tur\-a\-bi\-an, \break 1973) is one resource for 
the student writing a thesis.
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% figure page (see pp. 38 of the U of O Style Manual) %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                       %
% NOTE!!!! Setting up tables like this baby in TeX is not a real snap   %
% so if you plan on attempting it be prepared to be VERY patient while  %
% working it all out! You *CAN* do it, but it will take some work!      %
%                                                                       %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vglue .75truein
\vbox{\offinterlineskip 
\halign{#&#&#&#&#&#&#&#&#&#&#\cr

\hglue 1.0truept&\hglue 1.50truein&\hglue 1.0truept
&\hglue 0.25truein&
\hglue 1.0truept&\hglue 2.0truein&\hglue 1.0truept
&\hglue 0.25truein&
\hglue 1.0truept&\hglue 1.50truein&\hglue 1.0truept\cr

&&
&&
&\hrulefill\cr

\strut&&
&&
\hfil\vrule&\hfil Assistant Administrator\hfil&\vrule\cr

&&
&&
&\hrulefill\cr

&&
&&
&\hfil\vrule height25pt\hfil\cr

&&
&&
&\hrulefill\cr

\strut&&
&&
\hfil\vrule&\hfil Nursing Service Director\hfil&\vrule\cr

\strut&&
&&
\hfil\vrule&\hfil Specialty Units\hfil&\vrule\cr

&&
&&
&\hrulefill\cr

&&
&&
&\hfil\vrule height25pt\hfil\cr

&&
&&
&\hrulefill\cr

\strut&&
&&
\hfil\vrule&\hfil Surpervisor \hfil&\vrule\cr

&&
&&
&\hrulefill\cr

&&
&&
&\hfil\vrule height25pt\hfil\cr

&&
&&
&\hrulefill\cr

\strut&&
&&
\hfil\vrule&\hfil Nurse Unit Manager\hfil&\vrule\cr

&&
&&
&\hrulefill\cr

&&
&&
&\hfil\vrule height25pt\hfil\cr

&\hfill\hrulefill&\hrulefill&\hrulefill&\hrulefill&\hrulefill&\hrulefill&\hrulefill&
\hrulefill&\hrulefill\hfill\cr

&\hfil\vrule height25pt \hfil&
&&
&\hfil\vrule height25pt \hfil&
&&
&\hfil\vrule height 25pt \hfil\cr

&\hrulefill&
&&
&\hrulefill&
&&
&\hrulefill\cr

\strut\hfil\vrule&\hfil  Nurse       \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil Charge Nurse \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil  Nurse       \hfil&\vrule\hfil\cr
  
\strut\hfil\vrule&\hfil Facilitator \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil&\vrule\hfil&
&
\hfil\vrule&\hfil Facilitator \hfil&\vrule\hfil\cr

\strut\hfil\vrule&\hfil 7:00 A.M. to \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil 3:00 P.M. to \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil 11:00 P.M. to \hfil&\vrule\hfil\cr

\strut\hfil\vrule&\hfil 3:00 P.M. \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil 11:30 P.M. \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil 7:30 A.M. \hfil&\vrule\cr

&\hrulefill&
&&
&\hrulefill&
&&
&\hrulefill&\cr

&\hfil\vrule height25pt \hfil&
&&
&\hfil\vrule height25pt \hfil&
&&
&\hfil\vrule height 25pt \hfil\cr

&\hrulefill&
&&
&\hrulefill&
&&
&\hrulefill&\cr


\strut\hfil\vrule&\hfil Registered \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil Registered \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil Registered \hfil&\vrule\hfil\cr

\strut\hfil\vrule&\hfil Nurse \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil Nurse \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil Nurse \hfil&\vrule\hfil\cr

&\hrulefill&
&&
&\hrulefill&
&&
&\hrulefill&\cr

&\hfil\vrule height25pt\hfil&&&&\hfil\vrule height25pt\hfil\cr

&\hrulefill&
&&
&\hrulefill\cr

\strut\hfil\vrule&\hfil Unit Secretary \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil Unit Secretary\hfil&\vrule\hfil\cr

\strut\hfil\vrule&\hfil 8:00 A.M. to \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil 4:00 P.M. to \hfil&\vrule\hfil\cr

\strut\hfil\vrule&\hfil 4:00 P.M. \hfil&\vrule\hfil&
&
\hfil\vrule&\hfil 8:00 P.M. \hfil&\vrule\hfil\cr

&\hrulefill&&&&\hrulefill\cr}}
\vglue 1.0truein
\par\noindent
FIGURE 4. Nursing services authority structure for the neonatal intensive care unit.
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%
% table page (per p.39) %
%%%%%%%%%%%%%%%%%%%%%%%%%
\vglue 0.75truein
\baselineskip=\normalbaselineskip
\settabs 10 \columns
\centerline{TABLE 12. Analysis of Variance of Mean Initial After-Effect}
\centerline{for the Sighted Children, Ages 5 Through 16}
\bigskip\hrule width6.0truein \vskip2pt \hrule width6.0truein
\medskip
\+Source of&&&Sum of&&&Mean\cr
\+Variation&&&Squares&\hfill df&&Square&&F&p.\cr
\medskip\hrule width6.0truein\medskip
\+Between Groups&&&\enspace 2,222.22&\hfill 11&&202.02&&1.31&\hfill N.S.\cr
\smallskip
\+Within Groups&&&\hfill 19,822.12&\hfill 129&&153.55&\hfill\cr
\+&&&\hrulefill&\hfill\hrulefill&\cr
\+Total&&&\hfill 22,044.34&\hfill 140&\hfill\cr
\medskip
\hrule width6.0truein
\vglue 1.0truein
\centerline{TABLE 13. Methods Teachers Say They Would Employ Under New}
\centerline{Conditions as They Pertain to Increased Use of a}
\centerline{Variety of Learning Activities and Projects}
\bigskip\hrule width6.0truein \vskip2pt \hrule width6.0truein\medskip
\+&&&&&&&205 \hfil&&Rank&\cr
\+&&&&&&&Teachers\hfil&&Order of&\cr
\+Item&Reason&&&&&&Reporting\quad\quad\quad\quad\hfil&&Response&\cr
\+&&&&&&&\hrulefill&&\hrulefill&\cr
\+&&&&&&&No.\hfill \%&&Order\cr
\vskip1em\hrule width6.0truein \vskip1em
\+1.&More individual and small-group\cr
\+&projects&&&&&&17\hfill8.0&&\hfil\enspace 8\hfil&\cr
\medskip
\+2.&More activities with practical\cr
\+&application to subject matter\cr
\+&being taught&&&&&&13\hfill6.1&&\hfil 15 \hfil&\cr
\medskip  
\+3.&More creative activities&&&&&&12\hfill5.6&&\hfil 16\hfil&\cr
\medskip
\+4.&More freedom of choice of\cr 
\+&activities by pupils&&&&&&\enspace 7\hfill3.3&&\hfil 33\hfil&\cr
\medskip
\+5.&Enrichment activities for all\cr
\+&pupils in all areas of instruction&&&&&&\enspace 7\hfill3.3&&\hfil 34\hfil&\cr
\medskip   
\+6.&More child-centered project work&&&&&&\enspace 6\hfill2.8&&\hfil 38\hfil&\cr
\medskip
\+7.&More use of field trips&&&&&&\enspace 3\hfill1.4&&\hfil 70\hfil&\cr
\vskip 1em
\hrule width6.0truein \medskip\vfill\eject
\baselineskip=2\normalbaselineskip
\settabs 12 \columns
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% appendix page (see pp. 19 of the U of O Style Manual) %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vglue 2.75truein
\centerline{APPENDIX}
\vfill
\eject
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% bibliography page (see pp. 21 of the U of O Style Manual) %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vglue 2.75truein
\centerline{BIBLIOGRAPHY}
\vfill\eject
